Research Experience

### 2010-04-21 doc/thesis/thesis.tex
more work on the review

Increasing the spatial resolution of EPI imaging necessarily 
requires more time or faster magnetic field switching. Increasing
magnet switching rates though is difficult, because it can result in
more artifacts, or even lower signal to noise ratios. The result is
that at best FMRI is capable of 1.2 second temporal resolution. 
Additionally, the means that each voxel of the image will contain 
the sum of a large amount neurons, capillaries and veins. Thus, the
FMRI signal, which is sensitive to the chemical composition of 
materials, is summing up the composition of various types of tissue
in addition to the blood, whose composition is what we actually care about.
In particular, the presence of Deoxyhemoglobin, Hemoglobin whose
oxygen has been used by a metabolic process, has a larger magnetic
response compared to Oxygenated Hemoglobin. Thus, capillaries near
very active cells will typically have a higher Deoxyhemoglobin content,
and regions with lower activity will have a lower Deoxyhamoglobin content.
Unfortunately, as mentioned previously, blood is only a small part of
each voxel, which means that a single EPI image doesn't tell much about
Deoxyhemoglobin content. However assuming blood is the only thing changing
in the short term, percent difference from a baseline signal \emph{will}
tell us something. FMRI analysis is thus necessarily performed on the percent change
from the baseline. Luckily the assumption that tissue content does not change in
the short run is actually pretty good, although other facts can pollute
the baseline signal, as we will discuss later.

\subsection{BOLD Physiology}
The physiology that leads to the BOLD model has been well studied over the 
past decade and a half. While some uncertainty remains (for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>) generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events that locally 
alters the composition of the blood. Inactive neurons can be thought of
as a slingshot cocked and ready to fire. As soon as a signal moves up
the axon and causes the neuron to fire (and thus changes the state
of a membrane at that location), ions quickly move across the altered
membrane to compensate for a high charge imbalance between the interior
and exterior of the cell. This process, similar to allowing a stretched
rubber band to contract leaves the system at a lower energy state. This cannot
last though, because the neuron needs to be ready to fire again rather quickly.
To return the cell to a ready position the membrane becomes impermeable
to the ions again, and then begins pumping ions back into the cell. This
process takes a large amount of energy, whereas the actual firing takes
very little energy. Thus, after firing, glucose is burned, removing oxygen
from the blood and thus causing a dip in the ratio of oxygenated hemoglobin to 
de-oxygenated hemoglobin. This is the first potentially measurable effect
that MRI can see, although it usually lasts fewer than 2 seconds making it
difficult for FMRI to catch. As a result of the decreased amount of local oxygen,
the capillaries compensate by increasing blood flow to that region. Because,
of the quickly increased blood flow into local capillaries, the blood
volume increases in addition to the local oxygen content. This leads to a Windkessel
effect which further lags the normalization of oxygen content. The effect of
all this is an overshoot in the oxygen content above the initial level. After
the work of recharging is done, there may be a prolonged undershoot lasting
as much as 90 seconds \cite{spatial_undershoot}, though the reason for this
is debated \cite{origin_poststim}.

>>>>

Increasing the spatial resolution of EPI imaging necessarily 
requires more time or faster magnetic field switching. Increasing
magnet switching rates though is difficult, because it can result in
more artifacts, or even lower signal to noise ratios. The result is
that at \emph{best} FMRI is capable of 1 second temporal resolution. 
Additionally, the means that each voxel of the image will contain 
the sum of a large amount neurons, capillaries and veins. Thus, the
FMRI signal, which is sensitive to the chemical composition of 
materials, is summing up the composition of various types of tissue
in addition to the blood, whose composition is what we actually care about.
In particular, the presence of Deoxyhemoglobin, Hemoglobin whose
oxygen has been used by a metabolic process, has a decreased magnetic
response compared to Oxygenated Hemoglobin. Thus, capillaries near
very active cells will typically have a higher Deoxyhemoglobin content and
lower signal, and regions with lower activity will have a lower 
Deoxyhamoglobin content and thus higher signal.
Unfortunately, as mentioned previously, blood is only a small part of
each voxel, which means that a single EPI image doesn't tell much about
Deoxyhemoglobin content. However assuming blood is the only thing changing
in the short term, percent difference from a baseline signal \emph{will}
tell us something. FMRI analysis is thus necessarily performed on the percent change
from the baseline. Luckily the assumption that tissue content does not change in
the short run is actually pretty good, although other factors can pollute
the baseline signal, as we will discuss later.

\subsection{BOLD Physiology}
It is will known that Deoxyhemoglobin acts as a contrast agent in EPI imaging
\cite{Buxton1999}, \cite{WEISSKOFF1994}, \cite{Ogawa}, so then what causes
changes Deoxyhemoglobin in a voxel? Obviously increased metabolism will increase
Deoxyhemoglobin, but changes in blood volume in a voxel will also alter 
the MR signal depending on the ratio of Deoxyhemoglobin to hemoglobin. 
This was the impetus
for the ground breaking research \cite{Buxton1999} and \cite{Mandeville1999}.
These were the first two attempts to quantitatively account for the shape of the 
BOLD signal as a consequence of the lag between the cerebral blood flow (CBF) 
and the cerebral blood volume (CBV). In fact \cite{Buxton1999} went to far as
to show that a simple, well chosen blood flow waveform, along with a balloon 
model, could fully account for the BOLD signal. Although complex
changes in metabolism exist (\cite{Zheng2002}), most recent studies have
focused on their ability to explain the prolonged BOLD post stimulus 
undershoot \cite{Donahue2009}, \cite{Yacoub2006}. This is because \cite{Buxton2004}
and later \cite{Riera2004} showed that the primary signal may be accurately
be estimated by a simple flow-locked expression of the CMRO2. Comparison
between the more complicated models including a separate expression for CMRO2
and the simplified Buxton expression show little actual difference \cite{Deneux2006}.
In \cite{Buxton1999}, it was shown that small changes in the flow expression could
result in vastly different recovery times of the baseline BOLD signal, and
in \cite{Mandeville1999} a cause for delayed CBV recovery was proposed, based
on differences in compliance of capillaries. \cite{Behzadi2005} provided a simple,
complete model and explanation for delayed compliance of neural capillaries,
and those simple changes can result in a significant improvement in the bias error,
as shown in \cite{Deneux2006}. In the end, it may be impossible to refute the
possibility of a prolonged period of increased metabolic activity after activation,
however, much of the BOLD post stimulus undershoot may be explained by a 
prolonged CBV undershoot, as shown in \cite{Chan2009}. Because of the greatly
increased complexity of CMRO2 modeling, compared with the relative simplicity
of the Behzadi model, in most cases it is not worth using the more advanced CMRO2
model unless the metabolic rate is an end goal.

The physiology that leads to the BOLD signal has been well studied over the 
past decade and a half, but there is still some question about the origins
of certain signal features. In the seminal paper \cite{Buxton1999}, the original
Balloon model was proposed. Buxton et. al showed that all of the features
of the BOLD signal may be explained by lag between ... . While some uncertainty remains (for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>) generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events that locally 
alters the composition of the blood. Inactive neurons can be thought of
as a slingshot cocked and ready to fire. As soon as a signal moves up
the axon and causes the neuron to fire (and thus changes the state
of a membrane at that location), ions quickly move across the altered
membrane to compensate for a high charge imbalance between the interior
and exterior of the cell. This process, similar to allowing a stretched
rubber band to contract leaves the system at a lower energy state. This cannot
last though, because the neuron needs to be ready to fire again rather quickly.
To return the cell to a ready position the membrane becomes impermeable
to the ions again, and then begins pumping ions back into the cell. This
process takes a large amount of energy, whereas the actual firing takes
very little energy. Thus, after firing, glucose is burned, removing oxygen
from the blood and thus causing a dip in the ratio of oxygenated hemoglobin to 
de-oxygenated hemoglobin. This is the first potentially measurable effect
that MRI can see, although it usually lasts fewer than 2 seconds making it
difficult for FMRI to catch. As a result of the decreased amount of local oxygen,
the capillaries compensate by increasing blood flow to that region. Because,
of the quickly increased blood flow into local capillaries, the blood
volume increases in addition to the local oxygen content. This leads to a Windkessel
effect which further lags the normalization of oxygen content. The effect of
all this is an overshoot in the oxygen content above the initial level. After
the work of recharging is done, there may be a prolonged undershoot lasting
as much as 90 seconds \cite{spatial_undershoot}, though the reason for this
is debated \cite{origin_poststim}.
